Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper, hybrid and quant pred rationale in intro

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypes~\cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics~\cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough resampling routines applied to the subset of individuals 
with full information~\cite{Hastie2009}.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

@@

% Workings and advantages of the single step procedure.
Independently from each other,~\citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error's structure.

% -   Mention scarcity of data on single step prediction in plant populations.
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to plant hybrids, (2) explore the utility of single 
step predition with a quantitative predictor and (3) evaluate the impact of the
proporition of genotypes not covered by the incomplete predictor on predictive
ability.

>>>>

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypes~\cite{DeLosCampos2013,Garcia-Ruiz2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when all genotypes are equipped with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
estimated in the subset of individuals with full information.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

@@

% Workings and advantages of the single step procedure.
Independently from each other,~\citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error structure.

% -   Objectives (Hybrids)
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
While single step prediction has been applied to pig hybrids~\cite{Xiang2016}
this application is not equivalent to hybrid breeding in plants.
There, the maternal and the paternal crossing partner can be selected without
constraints whereas animal breeders select only males because of the limited
fertility of the female animals.
Technological advances in plant hybrid breeding such as the rapid generation
of completely homozygous lines~\cite{Wedzony2009} now enable breeders to
produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting these hybrid candidates with low relatedness to tested hybrids using
only information from their parents is currently a hot topic in plant breeding
\cite{Technow2014,Kadam2016,Westhues2017}.
Hence, our first objective was to examine whether the single step prediction 
framework can be transfered to hybrids in plants.

% -   Objectives (quantitative predictores)
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolites and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
This situation motivated our second objective, namely exploring the utility of
single step predition with a quantitative predictor.
Yet, these technologies are not fully mature and therefore incur higher costs
to breeders than pedigree and genomic data, putting their utility into question
despite their proven usefulness.
Our third objective is therefore to evaluate the impact of equipping only a
subset of all genotypes with both, the complete and the incomplete predictor on
predictive ability.
